---
title: "HW3 Peer Assessment"
output:
  html_document:
    df_print: paged
date: "`r format(Sys.time(), '%c %Z')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(tinytex.verbose = TRUE)

library(car)
```

# Part 1: Logistic Regression
# Background

The objective is to predict whether a passenger survived the Titanic disaster

## Data Description
1. **Survival**: (1 if the passenger survived, 0 if they did not).
2. **Pclass**: Numeric (1,2,3).
3. **Sex**: Categorical (male, female)
4. **Age**: Numeric.
5. **SibSp**: Number of siblings/spouses aboard. 
6. **Parch**: Number of parents/children aboard.
7. **Fare**：Numeric.

## Setup

You can import the data and set up the problem with the following R code:

```{r}
# Import the data
rawdata = read.csv("Titanic.csv", header=TRUE)

# Remove columns that we do not want
rawdata$X <- NULL
rawdata$PassengerId <- NULL
rawdata$Name <- NULL
rawdata$Ticket <- NULL
rawdata$Cabin <- NULL

# Set variables as categorical
rawdata$Pclass<-as.factor(rawdata$Pclass)
rawdata$Sex<-as.factor(rawdata$Sex)
rawdata$SibSp<-as.factor(rawdata$SibSp)
rawdata$Parch<-as.factor(rawdata$Parch)

# Print head of rawdata
head(rawdata)
```

**Note:** For all of the following questions, treat variables **Age** and **Fare** as quantitative variables and **Pclass**, **Sex**, **SibSp**,and **Parch** as categorical variables. Categorical variables have already been converted to factors in the starter code.


# Question 1: Fitting the Model - 9 pts

Use *Survived* as the response variable to fit a logistic regression model using logit as the link function with *Pclass* as the predictor. Call it **model1**.

**(a) 3 pts - Print the model1 summary table. What are the model parameters and estimates?**

```{r}

```

**(b) 6 pts - Calculate the odds ratio and interpret the estimated coefficient of *Pclass3* with respect to the log-odds of survival and the odds of survival.**

```{r}

```

Odds ratio: $\exp^{(-1.6390)}$ = 0.194

# Question 2: Statistical Inference - 7 pts 

**(a) 3 pts - Using model1, provide a 95% confidence interval for the coefficient of *Pclass3*.**

```{r}

```

**(b) 4 pts - According to the summary table of model1, is *Pclass3* statistically significant? Is the overall regression statistically significant at the 0.01 significance level?**

```{r}

```

# Question 3: Goodness of fit - 17 pts

**(a) 5 pts - Assess the goodness-of-fit of model1 using both deviance residuals and Pearson residuals. What do you conclude?**

```{r}

```

```{r}

```

**(b) 2 pts - Calculate the estimated dispersion parameter for this model. Is this an overdispersed model?**

```{r}

```

**(c) 2 pts - Using *Survived* as the response variable with *Pclass*, *Sex*, *SibSp*, and *Parch* as the predictors and logit as the link function, fit a logistic regression model and call it **model2**.**
```{r}

```

**(d) 3 pts - Using both deviance residuals and Pearson residuals to assess the goodness-of-fit hypothesis test. What do you conclude? **
```{r}

```

```{r}

```

**(e) 3 pts - Calculate the estimated dispersion parameter for this model. Compare model2 with model1 based on the p-values of the deviance residuals and Pearson residuals tests and the dispersion parameters of the two models. Which model is a better fit?**
```{r}

```

**(f) 2 pts - Please provide two methods that can potentially improve the model fit.**

# Question 4: Prediction - 7 pts

Suppose there is a passenger with the following characteristics:

1. **Pclass**: 3

2. **Sex**: male

3. **SibSp**: 2

4. **Parch**: 1

**(a) 4 pts - Predict the passenger's probability of surviving using model1 and model2.**

```{r}

```

```{r}

```

Model1 probability of survival = 0.251.

```{r}

```

**(b) 3 pts - Compare your predictions. i.e. which model is more reliable based on the analysis?**

# Part 2: Poisson Regression
# Background
Bicycle crossing counts for East River bridges crossing into Manhattan, New York City, and we want to focus on Brooklyn Bridge specifically.

## Data Description
1. **High.Temp...F.**: Numeric.
2. **Low.Temp...F.**: Numeric.
3. **Precipitation**: Numeric (range from 0-1).
4. **DayOfWeek**: Categorical (Sun, Mon, Tues, Wed, Thu, Fri, Sat).
5. **Brooklyn.Bridge**: number of bicyclists that cross Brooklyn Bridge in a day.

## Setup

You can import the data and set up the problem with the following R code:

```{r}
# Import data.
casedata <- read.csv("bicycle.csv", header=T)

casedata$Day <- as.Date(casedata$Day)
casedata$Precipitation <- as.numeric(casedata$Precipitation)

casedata$DayOfWeek <- weekdays(casedata$Day)
casedata$DayOfWeek <- as.factor(casedata$DayOfWeek)

casedata_brooklyn <- casedata[,c("High.Temp...F.", "Low.Temp...F.", "Precipitation", "Brooklyn.Bridge", "DayOfWeek")] 

head(casedata_brooklyn)
```

# Question 5: Fitting the Model - 7 pts

Using *Brooklyn.Bridge* as the response variable with *High.Temp...F.*, *Low.Temp...F.*, *Precipitation*, and *DayOfWeek* as the predictors, fit a poisson regression model and call it **poisson_model**.

**(a) 3 pts -  Print the poisson_model summary table. Interpret the estimated coefficient of *Precipitation* with respect to the log expected Brooklyn Bridge count.**
```{r}

```

**(b) 2 pts - Is the overall regression significant?**
```{r}

```

**(c) 2 pts - Perform a goodness-of-fit test on poisson_model using deviance residuals. What do you conclude?**
```{r}

```

# Question 6: Outlier detection - 13 pts

**(a) 2 pts - Create a Q-Q plot of the residuals for the poisson_model.**
```{r}

```

**(b) 4 pts - Using Cook's Distance with threshold = 1, how many outliers are identified in "poisson_model"? Remove the outliers from the dataset and create poisson_model2 with the new dataset.**
```{r}

```

```{r}

```

**(c) 4 pts - Perform the deviance residual test to check the goodness-of-fit of poisson_model2. Is poisson_model2 a good fit? Discuss the appropriateness of using Cook's Distance for poisson regression.**
```{r}

```

**(d) 3 pts - Use the alternative method for outlier detection discussed in lecture 13 on the first Poisson model ("poisson_model"). How many outliers are identified? What assumption is made in this approach?**

```{r}

```
